Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-06 talk/dls2012/paper.tex
some typos

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with a jump the the second part
(called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
Note that the peeled loop is not necessary the \emph{first} iteration of the
loop execution, it is general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration has happened.

@@

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy  & PyPy & GCC \\
 &         &       & no LP &      & -O3 \\
\hline
conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
\hline
conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
\hline
conv3x3(1000) & 233.54 & 125.40 & 0.57 & 0.27 & 0.25 \\
\hline
conv3x3(3) & 234.45 & 126.28 & 0.60 & 0.31 & 0.28 \\
\hline
conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
\hline
conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
\hline
dilate3x3(1000) & 232.51 & 125.85 & 3.89 & 3.69 & 0.25 \\
\hline
sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
\hline
\hline
Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
  $10^6$ and matrixes of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrixes, only the number of rows are
  specified.} 
\end{figure}

@@

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developped the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
\cite{pall_luajit_2009}

>>>>

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with a jump to the second part
(called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintain the SSA-property.
Note that the peeled loop is not necessary the \emph{first} iteration of the
loop execution, it is general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration has happened.

@@

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy  & PyPy & GCC \\
 &         &       & no LP &      & -O3 \\
\hline
conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
\hline
conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
\hline
conv3x3(1000) & 233.54 & 125.40 & 0.57 & 0.27 & 0.25 \\
\hline
conv3x3(3) & 234.45 & 126.28 & 0.60 & 0.31 & 0.28 \\
\hline
conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
\hline
conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
\hline
dilate3x3(1000) & 232.51 & 125.85 & 3.89 & 3.69 & 0.25 \\
\hline
sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
\hline
\hline
Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
  $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrices, only the number of rows are
  specified.} 
\end{figure}

@@

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developed the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
\cite{pall_luajit_2009}
